---
title: "PS2"
author: "Jenny Jia"
date: "2026-03-03"
output: pdf_document
---
GitHub Repository URL: https://github.com/jennyjia319/gov-51-ps2

# 1 Part 1: The Original Study
## 1.1 Exploring the Experiment
### 1.1.1 Load Packages and Data
```{r}
library(tidyverse)
gay <- read.csv("data/raw/gay.csv")
```
### Q1.1
```{r}
nrow(gay) # look at total observations
n_distinct(gay$study) # look at unique values
unique(gay$treatment) #list treatment conditions
```
There are 69,592 total observations in the dataset. There are 2 unique values of study. The treatment conditions are: "no contact", "recycling script by gay canvasser", "same-sex marriage script by gay canvasser", "recycling script by straight canvasser", and "same-sex marriage script by straight canvasser". 

### Q1.2
```{r}
study1wave1 <- gay |> # create new dataset that only includes study 1 and wave 1
  filter(study == 1, wave == 1)
mean <- study1wave1 |> 
  group_by(treatment) |> 
  summarize(mean_ssm = mean(ssm, na.rm = TRUE)) |> 
  arrange(desc(mean_ssm)) # arrange for easier viewing 
mean # to show the highest means
```
Within study 1 wave 1, treatment group "Recycling Script by Gay Canvasser" shows the highest average support, with a mean of 3.13. 

### Q1.3
```{r}
gay |>
  filter(study == 1, wave == 1) |>
  group_by(treatment) |> #separate by treatment group
  summarize(sd_ssm = sd(ssm, na.rm = TRUE)) # SD calculation
```
The standard deviations are:
1.68 for No Contact
1.70 for Recycling Script by Gay Canvasser
1.68 for Recycling Script by Straight Canvasser
1.67 for Same-Sex Marriage Script by Gay Canvasser
1.68 for Same-Sex Marriage Script by Straight Canvasser. 
These standard deviations fall within the narrow range of 1.67-1.70. What this indicates is that the variability in attitudes is comparable across the 5 groups. This is positive, and reflects the assumption that randomization creates balanced groups. In the case that a group had a noticeably larger or smaller standard deviation, it would complicate how we can determine the fact that the treatment is solely responsible for the differences in means. 

### Q1.4
```{r}
study1wave1_means <- gay |>
  filter(study == 1, wave == 1) |>   
  group_by(treatment) |>            
  summarize(mean_ssm = mean(ssm, na.rm = TRUE) 
  )
study1wave1_means
```
```{r, fig.width=13, fig.height=5} 
#changed dimensions because default dimensions were too narrow
ggplot(study1wave1_means,
       aes(x = reorder(treatment, -mean_ssm), y = mean_ssm)) +
  geom_col() +
  labs(
    title = "Mean Same-Sex Marriage Support: Study 1, Wave 1",
    x = "Treatment Group",
    y = "Mean SSM (1–5)"
  )
```
In general, the means are so similar that it's hard to see the differences on the graph. 

## 1.2 Checking Balance and Comparing Outcomes
### Q1.5 
```{r}
study1wave1_check <- gay |>
  filter(study == 1, wave == 1,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact")
  )
stats <- study1wave1_check |>
  group_by(treatment) |>
  summarize(
    mean_ssm = mean(ssm, na.rm = TRUE),
    sd_ssm = sd(ssm, na.rm = TRUE)
  )

stats
```
```{r}
mean_gayscript <- stats$mean_ssm[
  stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

mean_nocontact <- stats$mean_ssm[
  stats$treatment == "No Contact"
]

sd_gayscript <- stats$sd_ssm[
  stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

sd_nocontact <- stats$sd_ssm[
  stats$treatment == "No Contact"
]

SDM <- (mean_gayscript - mean_nocontact) /
  sqrt((sd_gayscript^2 + sd_nocontact^2) / 2)

SDM
```
A standardized difference of -0.01 is achieved when rounded. This is very close to 0 and falls well below the 0.25 threshold for balance. What this means is that, at a baseline, Same-sex Marriage Script by Gay Canvasser group and the No Contact group are nearly identical in terms od attitudes. Thus, randomization created balanced groups. 

### Q1.6
```{r}
study1wave2_check <- gay |>
  filter(study == 1, wave == 2,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact")
  )
wave2_stats <- study1wave2_check |>
  group_by(treatment) |>
  summarize(
    mean_ssm = mean(ssm, na.rm = TRUE)
  )
wave2_stats
```
```{r}
mean_gayscript_w2 <- wave2_stats$mean_ssm[
  wave2_stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"]

mean_nocontact_w2 <- wave2_stats$mean_ssm[
  wave2_stats$treatment == "No Contact"]

SDM2 <- mean_gayscript_w2 - mean_nocontact_w2

SDM2
```
Now looking at Study 1 Wave 2, the mean ssm for No Contact is 3.039615 (3.04 when rounded) and 3.139489 (3.14 when rounded) for Same-sex Marriage Script by Gay Canvasser. The difference in means is 0.09987463 points, or 0.10 when rounded. The difference at wave 2 must suggest, then, that contact with a gay canvasser script moderately shifted attitudes on same-sex marriage relative to the no contact group.
