---
title: "PS2"
author: "Jenny Jia"
date: "2026-03-03"
output: pdf_document
---
GitHub Repository URL: https://github.com/jennyjia319/gov-51-ps2

# 1 Part 1: The Original Study
## 1.1 Exploring the Experiment
### 1.1.1 Load Packages and Data
```{r}
library(tidyverse)
gay <- read.csv("data/raw/gay.csv")
```
### Q1.1
```{r}
nrow(gay) # look at total observations
n_distinct(gay$study) # look at unique values
unique(gay$treatment) #list treatment conditions
```
There are 69,592 total observations in the dataset. There are 2 unique values of study. The treatment conditions are: "no contact", "recycling script by gay canvasser", "same-sex marriage script by gay canvasser", "recycling script by straight canvasser", and "same-sex marriage script by straight canvasser". 

### Q1.2
```{r}
study1wave1 <- gay |> # create new dataset that only includes study 1 and wave 1
  filter(study == 1, wave == 1)
mean <- study1wave1 |> 
  group_by(treatment) |> 
  summarize(mean_ssm = mean(ssm, na.rm = TRUE)) |> 
  arrange(desc(mean_ssm)) # arrange for easier viewing 
mean # to show the highest means
```
Within study 1 wave 1, treatment group "Recycling Script by Gay Canvasser" shows the highest average support, with a mean of 3.13. 

### Q1.3
```{r}
gay |>
  filter(study == 1, wave == 1) |>
  group_by(treatment) |> #separate by treatment group
  summarize(sd_ssm = sd(ssm, na.rm = TRUE)) # SD calculation
```
The standard deviations are:
1.68 for No Contact
1.70 for Recycling Script by Gay Canvasser
1.68 for Recycling Script by Straight Canvasser
1.67 for Same-Sex Marriage Script by Gay Canvasser
1.68 for Same-Sex Marriage Script by Straight Canvasser. 
These standard deviations fall within the narrow range of 1.67-1.70. What this indicates is that the variability in attitudes is comparable across the 5 groups. This is positive, and reflects the assumption that randomization creates balanced groups. In the case that a group had a noticeably larger or smaller standard deviation, it would complicate how we can determine the fact that the treatment is solely responsible for the differences in means. 

### Q1.4
```{r}
study1wave1_means <- gay |>
  filter(study == 1, wave == 1) |>   
  group_by(treatment) |>            
  summarize(mean_ssm = mean(ssm, na.rm = TRUE) 
  )
study1wave1_means
```
```{r, fig.width=13, fig.height=5} 
#changed dimensions because default dimensions were too narrow
ggplot(study1wave1_means,
       aes(x = reorder(treatment, -mean_ssm), y = mean_ssm)) +
  geom_col() +
  labs(
    title = "Mean Same-Sex Marriage Support: Study 1, Wave 1",
    x = "Treatment Group",
    y = "Mean SSM (1–5)"
  )
```
In general, the means are so similar that it's hard to see the differences on the graph. 

## 1.2 Checking Balance and Comparing Outcomes
### Q1.5 
```{r}
study1wave1_check <- gay |>
  filter(study == 1, wave == 1,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact")
  )
stats <- study1wave1_check |>
  group_by(treatment) |>
  summarize(
    mean_ssm = mean(ssm, na.rm = TRUE),
    sd_ssm = sd(ssm, na.rm = TRUE)
  )

stats
```
```{r}
mean_gayscript <- stats$mean_ssm[
  stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

mean_nocontact <- stats$mean_ssm[
  stats$treatment == "No Contact"
]

sd_gayscript <- stats$sd_ssm[
  stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

sd_nocontact <- stats$sd_ssm[
  stats$treatment == "No Contact"
]

SDM <- (mean_gayscript - mean_nocontact) /
  sqrt((sd_gayscript^2 + sd_nocontact^2) / 2)

SDM
```
A standardized difference of -0.01 is achieved when rounded. This is very close to 0 and falls well below the 0.25 threshold for balance. What this means is that, at a baseline, Same-sex Marriage Script by Gay Canvasser group and the No Contact group are nearly identical in terms od attitudes. Thus, randomization created balanced groups. 

### Q1.6
```{r}
study1wave2_check <- gay |>
  filter(study == 1, wave == 2,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact")
  )
wave2_stats <- study1wave2_check |>
  group_by(treatment) |>
  summarize(
    mean_ssm = mean(ssm, na.rm = TRUE)
  )
wave2_stats
```
```{r}
mean_gayscript_w2 <- wave2_stats$mean_ssm[
  wave2_stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"]

mean_nocontact_w2 <- wave2_stats$mean_ssm[
  wave2_stats$treatment == "No Contact"]

SDM2 <- mean_gayscript_w2 - mean_nocontact_w2

SDM2
```
Now looking at Study 1 Wave 2, the mean ssm for No Contact is 3.039615 (3.04 when rounded) and 3.139489 (3.14 when rounded) for Same-sex Marriage Script by Gay Canvasser. The difference in means is 0.09987463 points, or 0.10 when rounded. The difference at wave 2 must suggest, then, that contact with a gay canvasser script moderately shifted attitudes on same-sex marriage relative to the no contact group.
# 2 Part 2: The Forensic Discovery
## 2.1 The Baseline Distribution
### Q2.1
```{r}
gay_r <- read_csv("data/raw/gayreshaped.csv") 
ccap <- read_csv("data/raw/ccap2012.csv")

nrow(gay_r) # number of observations
nrow(ccap)

names(gay_r) # names of columns
names(ccap)
```
There are 11,948 observations in the gayreshaped dataset, and 43,998 observations in the ccap dataset. In the gayreshaped dataset, there are 6 variables: "treatment", "therm1", "therm2", therm3", "therm4". The therm variables measure thermometer scores from 0-100 toward gay men and lesbians. The ccap dataset has 6 variables: "..1", "caseid", "gaytherm", where gaytherm measures thermometer attitudes toward gay men and lesbians. 

### Q2.2
```{r}
gay_baseline <- gay_r |>
  filter(study == 1)

gay_stats <- gay_baseline |>
  summarize(
    mean = mean(therm1, na.rm = TRUE),
    median = median(therm1, na.rm = TRUE),
    sd = sd(therm1, na.rm = TRUE)
  )

ccap_stats <- ccap |>
  summarize(
    mean = mean(gaytherm, na.rm = TRUE),
    median = median(gaytherm, na.rm = TRUE),
    sd = sd(gaytherm, na.rm = TRUE)
  )

comparison <- rbind( #combine both for side by side comparison 
  cbind(dataset = "Gay Study", gay_stats),
  cbind(dataset = "CCAP", ccap_stats)
)

comparison
```
The scores in the Study 1 sample have a mean of 58.38 (when rounded), median of 52, and standard deviation of 28.50 (when rounded). In the 2012 CCAP dataset, the mean is 58.71 (when rounded), median is 54, and standard deviation is 29.37. These numbers are noticeably similar. Given the context of the study, this raises questions on whether the data has been fabricated to appear smoother. 

### Q2.3
```{r}
gay_histogram <- gay_r |>
  filter(study == 1) |>
  select(therm1) |>
  mutate(source = "LaCour Study 1 Baseline",
         therm = therm1) |>
  select(source, therm)

# CCAP thermometer (remove missing values)
ccap_histogram <- ccap |>
  filter(!is.na(gaytherm)) |>
  mutate(source = "CCAP 2012",
         therm = gaytherm) |>
  select(source, therm)

combined_histogram <- rbind(gay_histogram, ccap_histogram) #combined into one dataset

ggplot(combined_histogram, aes(x = therm)) + # side by side historgram creation
  geom_histogram(breaks = seq(0, 100, by = 5), color = "lightpink") +
  facet_wrap(~ source) +
  labs(
    title = "Comparison of Thermometer Distributions",
    x = "Feeling Score Toward Gay Men and Lesbians (0–100)",
    y = "Number of Respondents"
  )
```
Shape-wise, both distributions look generally similar, with more concentration in respondents around the scores of 50 and 100 in both datasets. Because they appear so similar, this may be a red flag, as both these samples were supposed to have been collected independently from each other, so it is rather suspicious how they could appear so similar. 

## 3.2 Checking Balance and Comparing Outcomes in Study 2
### Q3.3 
```{r}
study2_check <- gay_r |> # now we do study 2
  filter(
    study == 2,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact"
    )
  )

study2_stats <- study2_check |> #mean and sd calculations
  group_by(treatment) |>
  summarize(
    mean_therm1 = mean(therm1, na.rm = TRUE),
    sd_therm1 = sd(therm1, na.rm = TRUE)
  )

study2_stats
```
```{r}
mean_gayscript_study2 <- study2_stats$mean_therm1[
  study2_stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

mean_nocontact_study2 <- study2_stats$mean_therm1[
  study2_stats$treatment == "No Contact"
]

sd_gayscript_study2 <- study2_stats$sd_therm1[
  study2_stats$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

sd_nocontact_study2 <- study2_stats$sd_therm1[
  study2_stats$treatment == "No Contact"
]

SDMstudy2 <- (mean_gayscript_study2 - mean_nocontact_study2) /
  sqrt((sd_gayscript_study2^2 + sd_nocontact_study2^2) / 2)

SDMstudy2
```
In study 2 at baseline, the mean and standard deviation of the No Contact group was 57.89 (when rounded) and 28.70 (when rounded). The Same-sex Marriage Script by Gay Canvasser group had a mean of 59.35 (when rounded) and standard deviation of 28.42 (when rounded). The standardized difference in means is 0.05128243 (0.05 when rounded). This is far below the 0.25 threshold, which indicates that the groups look well-balanced. 

### Q3.4
```{r}
study2_postconversation <- gay_r |> 
  filter(
    study == 2,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact"
    )
  ) |>
  group_by(treatment) |>
  summarize(
    mean_therm2 = mean(therm2, na.rm = TRUE)
  )

study2_postconversation
```
```{r}
mean_gay_postconversation <- study2_postconversation$mean_therm2[ # difference
  study2_postconversation$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

mean_no_postconversation <- study2_postconversation$mean_therm2[
  study2_postconversation$treatment == "No Contact"
]

comparison_study2 <- mean_gay_postconversation - mean_no_postconversation

comparison_study2
```
In Study 2, the mean post-conversation thermometer score is 61.20 for the Same-Sex Marriage Script by Gay Canvasser group and 57.64 for the No Contact group (both when rounded). The difference in means is 3.57 points. Since the groups were balanced at baseline, this difference suggests that the canvassing treatment did increase favorable attitudes toward gay men and lesbians.

## 3.3 Uncertainty Around the Difference
### Q3.5
```{r}
study2_post_summary <- gay_r |> # filter to study 2
  filter(study == 2, treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact")
  ) |>
  group_by(treatment) |> # calculations for values 
  summarize(
    mean_therm2_study2 = mean(therm2, na.rm = TRUE),
    variance_therm2_study2 = var(therm2, na.rm = TRUE),
    n_therm2_study2 = sum(!is.na(therm2)) #sample size
  )

study2_post_summary
```
```{r}
mean_gay_therm2_study2 <- study2_post_summary$mean_therm2_study2[ # means for each group
  study2_post_summary$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

mean_no_therm2_study2 <- study2_post_summary$mean_therm2_study2[
  study2_post_summary$treatment == "No Contact"
]

variance_gay_therm2_study2 <- study2_post_summary$variance_therm2_study2[ # variances for each group
  study2_post_summary$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

variance_no_therm2_study2 <- study2_post_summary$variance_therm2_study2[
  study2_post_summary$treatment == "No Contact"
]

n_gay_therm2_study2 <- study2_post_summary$n_therm2_study2[ #sample sizes for each group
  study2_post_summary$treatment == "Same-Sex Marriage Script by Gay Canvasser"
]

n_no_therm2_study2 <- study2_post_summary$n_therm2_study2[
  study2_post_summary$treatment == "No Contact"
]

# standard error calculations

SE_diff_therm2_study2 <- sqrt(
  (variance_gay_therm2_study2 / n_gay_therm2_study2) +
  (variance_no_therm2_study2 / n_no_therm2_study2)
)

SE_diff_therm2_study2
```
```{r}
diff_means_therm2_study2 <- 
  mean_gay_therm2_study2 - mean_no_therm2_study2 # must find difference in means for CI calculation

lower <- diff_means_therm2_study2 - 1.96 * SE_diff_therm2_study2 # confidence interval calculations

upper <- diff_means_therm2_study2 + 1.96 * SE_diff_therm2_study2

c(lower, upper)
```
The standard error is 1.24 when rounded, and the 95% confidence interval is 1.13, 6.00 when rounded. This interval does NOT include 0. Thus, we can conclude that the difference between Gay Canvasser and No Contact groups is statistically significant, at the 5% level. We can say that the act of canvassing increased feeling scores by approximately 1-6 points in study 2. 

## 3.4 Tracking Attitude Over Time
### 3.6 
```{r}
study2_control <- gay_r |> # filter for study 2 only
  filter(
    study == 2,
    treatment == "No Contact"
  )

correlation12 <- cor(study2_control$therm1, # correlation between wave1 and wave2
               study2_control$therm2,
               use = "complete.obs")

correlation13 <- cor(study2_control$therm1, # correlation between wave1 and wave3
               study2_control$therm3,
               use = "complete.obs")

correlation14 <- cor(study2_control$therm1, # correlation between wave1 and wave4
               study2_control$therm4,
               use = "complete.obs")

correlation12
correlation13
correlation14
```
The correlation between therm1 and therm2 is 0.97, between therm1 and therm3 is 0.96, and between therm1 and therm4 is 0.97 (all rounded values). These correlations are extremely high and do not show a clear pattern of decay over time, as the correlation with therm4 is nearly the same as with therm2 In real survey data, we would typically expect correlations to decline as waves become more distant due to natural attitude change, so this stable pattern does raise concerns about the authenticity of the data.

### Q3.7
Study 1 and 2 differ and size and how realistic the results are. In study 2, the post-period treatment effect is 3.57 on the 0-100 thermometer scale, with a 95% confidence interval of 1.13, 6.00. This means that there is a statistically significant but uncertain effect. In contrast, study 1 showed only a 0.10 difference on the 1-5 scale at wave 2, despite the data appearing suspiciously smooth. Usually, real survey data would contain potential variation and errors, so treatment effects should be rather modest. It is important that real treatment effects are smaller and messier because peoples' attitudes are constantly changing. It's actually quite cool to see how we can pick out fabrications in data through analyses tools as simple as histograms and mean calculations. Sometimes, it takes visualization or direct side by side comparison of data to debunk patterns that look too "perfect" to be genuine. 